---
title: "Environment Modules"
output: rmarkdown::github_document
vignette: >
  %\VignetteIndexEntry{readme}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Note: knit this to the repo readme file using:
# rmarkdown::render("vignettes/readme.Rmd", output_format = "github_document", output_dir = getwd())

process_ansi_string = function(x, options) {
  # Truncate the output if it's too long.
  # Adapted from: https://bookdown.org/yihui/rmarkdown-cookbook/hook-truncate.html
  if (!is.null(n <- options$out.lines)) {
    attribs = attributes(x)
    if (length(x) > n) {
      x = c(head(x, n), "....\n")
    }
    x = paste(x, collapse = "\n")
    attributes(x) = attribs
  }
  
  # Style it as HTML
  # Adapted from: https://cli.r-lib.org/reference/ansi_html.html#ref-examples
  hcode = paste(ansi_html(x), collapse = "\n")
  
  # Strip CSS for github
  if (knitr::pandoc_to("gfm")){
    head = htmltools::tags$head()
  }
  else {
    head = htmltools::tags$head(htmltools::tags$style(paste(format(ansi_html_style()), collapse = "\n")))
  }
  htmltools::tagList(
    head,
    htmltools::tags$pre(htmltools::HTML(hcode))
  ) |>
    htmltools::doRenderTags() |>
    knitr::asis_output()
}

knit_print.cliMessage = process_ansi_string

knit_print.ansi_string = process_ansi_string

knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

#knitr::opts_chunk$set(message = process_ansi_string)
```

The EnvironmentModules package is an R front-end for the [Environment Modules](https://modules.readthedocs.io/en/latest/index.html) project, which offers a way to load and unload packages, and is commonly used on shared HPC (High Performance Computing) environments.

```{r setup}
library(EnvironmentModules)
```

```{r, include=FALSE}
module_purge()
```


## Listing Modules

You can list the available modules:
```{r out.lines=20}
module_avail()
```

You can filter to only modules starting with some prefix:

```{r}
module_avail("python")
```

## Loading and Unloading Modules

Load modules using `module_load()`!

```{r}
module_load("python/3")
```

If you've already loaded a module, you will be notified:

```{r}
module_load("python/3")
```

You can unload a module using the corresponding `module_unload()`:
```{r}
module_unload("python/3")
```

## Listing Loaded Modules

You can list modules that have been already loaded with `module_list()`:

```{r}
module_load("python/3")
module_list()
```
## Linking Shared Libraries

Although `module_load()` will add a module's libraries to your `LD_LIBRARY_PATH`, which will allow you to compile packages against them, R will not necessarily be able to actually run packages that use these libraries.

For example, let's say we want to use the `hdf5r` package, which depends on the `hdf5` library package. We don't have `hdf5` loaded, so it will fail:

```{r}
#install.packages("hdf5r", quiet = TRUE)
```

We can resolve this by loading the appropriate module. Firstly, we need to find out what the module is called:

```{r}
module_avail("hdf5")
```
Then loading the module:

```{r}
module_load("hdf5/1.12.2")
```

Now let's try again:

```{r}
#install.packages("hdf5r", quiet = TRUE)
```

Finally, we can load the package itself... or can we?

```{r, error=TRUE}
library(hdf5r)
```
As alluded to above, R doesn't actually load every new library that becomes available in the environment.
Rather it uses "load-time linking", which is where the libraries it links to are fixed at the time you start R.
Now, you can resolve the above error by closing R, loading the module, and then restarting R, but this can be quite annoying.

Fortunately, R also supports "run-time linking", which you can enable by loading a module with the `link_libs=TRUE` argument:

```{r}
module_unload("hdf5/1.12.2")
module_load("hdf5/1.12.2", link_libs=TRUE)
```

```{r}
library(hdf5r)
```

It worked!

If you are interested in the theory underlying what is happening here, you might find this explanation of linking on Linux helpful: https://techblog.rosedu.org/library-management.html.


## Other Commands

This package also supports the following.

`module_swap()` unloads the first module and loads the second:

```{r}
module_swap("hdf5/1.12.2", "hdf5/1.10.5")
```

`module_purge()` unloads all modules:

```{r}
module_purge()
```

## Advanced Commands

If you want to use a feature of Environment Modules that does not (yet) have a dedicated function, you can use the advanced functions `get_module_code` or `get_module_output`.

If you want to run a command that doesn't edit the modules, but only shows output, run `get_module_output`

```{r}
get_module_output(c("show", "hdf5"))
```

If the module does edit the modules, use `get_module_code()`, and evaluate the result:

```{r}
get_module_code("reload") |> eval()
```

## FAQ

> Environment Modules [supports loading modules in R since version 4.0.0!](https://modules.readthedocs.io/en/latest/index.html) What is the point of this package?

This is true, and actually this package is just a wrapper around that core functionality. You *need* Modules version 4.0.0 or above to run this package. The main advantages of using this wrapper are the nicer function interfaces, and the ability to automatically link R to newly loaded libraries.

> I'm getting the error "Could not detect an Environment Modules installation", but I know that Environment Modules is installed on my system. What do I do?

First, in a bash terminal where the `module` command works correctly, run `which modulecmd`.
Then in R, run `Sys.setenv(MODULES_CMD="</path/to/modulecmd>"), putting the output from the previous step as the argument value.
Then everything should work!
